Figures for Paper

### 2021-07-06 paper_figures.Rmd
STR genotype + more metrics graphs

## Effect of library size on CE discovery

@@

```{r rin discovery}
## effect of RIN?
clean_data_table %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  filter(!is.na(rin)) %>%
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = unc13a_detected,y = rin )) + 
  geom_point( colour = "firebrick", size = 0.5, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
   geom_boxplot(fill = NA, colour = "black", outlier.colour = NA) +
  labs(title = "Sample RIN and UNC13A CE detection ",subtitle = "Wilcox test") + 
  facet_wrap(~disease_group2 + tissue) + 
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center" ) +
  labs(y = "RIN", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()

@@

```{r unc13a ce and UNC13B_TPM}
## effect of RIN?
clean_data_table %>%
  filter(disease_tissue == T) %>% 
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  add_count(tissue_clean,disease_group2) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n ({n})")) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = plot_name, fill = unc13a_detected,y = UNC13B_TPM )) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.colour = NA) +
  labs(title = "UNC13A CE detection and UNC13B TPM",subtitle = "Wilcox test") + 
  facet_wrap(~ tissue_clean,scales = 'free') + 
  ggpubr::stat_compare_means(label = "p.signif",) +
  labs(y = "UNC13B TPM", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()
```

@@

```{r}
conversion_table <- data.frame(short = c("Ast","End", "Mic", "Ex", "Oli", "Per" ),
                               long = c("Astrocytes", "Endothelial", "Microglia", "Neurons", "Oligos", "Pericytes")
  )

cell_types = fread("/Users/annaleigh/Documents/GitHub/unc13a_cryptic_splicing/data/MuSiC_residual_results.tsv")

cell_types %>% 
  left_join(clean_data_table, by = 'sample') %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-")) %>% 
  filter(!is.na(inclusion_reads)) %>% 
  select(sample,cell,deconv, tissue_clean,unc13a_detected) %>% 
  unique() %>% 
  ggplot() + 
  geom_boxplot(aes(x = unc13a_detected, y = deconv, fill = cell)) + 
  facet_wrap(~ tissue_clean,scales = 'free')

@@

in_psi = rs12608932_comparisons %>%
  ggplot(aes(y = unc13a_cryptic_leaf_psi, x = detection_name, fill = rs12608932)) +
  geom_boxplot() +
  geom_jitter( height = 0) +
  ylab(element_blank()) +
  xlab(element_blank()) +
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ggpubr::theme_pubr() +
  theme(legend.position = "none") +
  theme(text = element_text(size = 24)) +
  stat_compare_means() +
  stat_compare_means(comparisons = comp_for_gg,label = "p.signif") +
    scale_y_continuous(trans='log10',labels = scales::percent_format(accuracy = 1L))  +
  ggtitle("intronic SNP (rs12973192)")
```

@@

@@

@@

```{r message=FALSE, warning=FALSE}
read_length_filter <- 29  # length of reads of interest
gencode_v29_info <- read_tsv(paste0(here(), "/data/longest_proteincoding_transcript_hs_details.txt"))
ribo_df <- read_tsv(paste0(here(), "/data/riboseq/sh_tdpb.Aligned.toTranscriptome.out.bam.bed.gz"),
                    col_names = c("transcript_id", "start0", "end", "strand")) %>%
    inner_join(gencode_v29_info, by = "transcript_id") %>%
    dplyr::filter(end-start0 == read_length_filter, strand == "+") %>%
    mutate(distance_from_start = start0-cds_start) %>%
    group_by(distance_from_start) %>%
    mutate(n_this_dist = n()) %>%
    dplyr::select(distance_from_start, n_this_dist) %>%
    unique()

@@

>>>>

## Effect of seqeuncer (read length) size on CE discovery

```{r unc detection by sequencer  }
# library size 
#llumina HiSeq 2500 (125 bp paired end) or an Illumina NovaSeq (100 bp paired end). 
unc13a_sequencing_platform <- clean_data_table %>%
  filter(tissue_clean %in% c("Frontal_Cortex", 
                             "Lumbar_Spinal_Cord", 
                             "Cervical_Spinal_Cord", 
                             "Motor_Cortex", 
                             "Temporal_Cortex")) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-")) %>% 
  mutate(tissue = as.character(tissue)) %>%
  filter(sequencing_platform != "") %>% 
  filter(disease_group2 %in% c("ALS-TDP","FTLD-TDP"))

quick_fisher = function(tbl){

  if(nrow(tbl) <2){
    return(NA)
  }else{
    comparison_table = table(unc13a = tbl$unc13a_detected, 
                             sequencing_platform = tbl$sequencing_platform)
    p_value = fisher.test(comparison_table)$p.value

    return(p_value)
  }
}

unc13a_data_fisher = unc13a_sequencing_platform %>% 
  filter(tissue_clean != "Thoracic_Spinal_Cord") %>% 
  split(.,list(unique(unc13a_sequencing_platform$tissue_clean), unique(unc13a_sequencing_platform$disease_group2))) %>% 
  purrr::map_df( quick_fisher, .id = "tissue_clean" ) %>% 
  melt() %>% 
  select(variable,value) %>% 
  separate(variable,sep = "\\.", into = c("tissue_clean","disease_group2"))

unc13a_sequencing_platform_pct = unc13a_sequencing_platform %>%
  group_by(disease_group2,tissue_clean, sequencing_platform, unc13a_detected) %>% tally() %>%
  spread(key = unc13a_detected, value = n,fill = 0) %>%
  mutate(detection = `+` / ( `-` + `+`) ) %>% 
  mutate(total =  ( `-` + `+`) ) %>% 
  mutate(plot_name = glue::glue("{sequencing_platform} \n  ({total})")) %>% 
  left_join(unc13a_data_fisher)

unc13a_sequencing_platform_pct %>%
  ggplot(aes(x = plot_name, y = detection )) + 
  geom_col(fill = "firebrick") +
  labs(title = "UNC13A CE detection and sequencing platform", y = "UNC13A CE detected") + 
  facet_wrap(~disease_group2 + tissue_clean,scales = 'free_x') +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(aes(x = 1.5, y = 1.05, label = paste0("p = ", signif(value,digits = 2) ) )) +
  theme(legend.text = element_text(size = 15))
 # scale_y_continuous(expand = c(0,0) ) 
```
## Effect of library size on CE discovery

@@

```{r rin discovery}
## effect of RIN?
clean_data_table %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  filter(!is.na(rin)) %>%
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
    mutate(disease_group2 = fct_relevel(disease_group2, "Control")) %>% 
  ggplot(aes(x = disease_group2,y = rin, fill = unc13a_detected)) + 
  geom_point( colour = "firebrick", size = 0.5, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
   geom_boxplot(colour = "black", outlier.colour = NA) +
  labs(title = "Sample RIN and UNC13A CE detection ") + 
  facet_wrap(~ tissue, scales = 'free_x') + 
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center") +
  labs(y = "RIN", x = element_blank(), fill = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()

@@

```{r unc13a ce and UNC13B_TPM}
## effect of RIN?
clean_data_table %>%
  filter(disease_tissue == T) %>% 
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  add_count(tissue_clean,disease_group2) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n ({n})")) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = plot_name, fill = unc13a_detected,y = UNC13B_TPM )) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.colour = NA) +
  labs(title = "UNC13A CE detection and UNC13B TPM",subtitle = "Wilcox test") + 
  facet_wrap(~ tissue_clean,scales = 'free') + 
  ggpubr::stat_compare_means(label = "p.signif",) +
  labs(y = "UNC13B TPM", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr() +
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center")
```

@@

```{r celltype and detection}

cell_types = fread("/Users/annaleigh/Documents/GitHub/unc13a_cryptic_splicing/data/All_1917_samples_Darmanis_dtangle_deconv.tsv")

cell_types %>% 
  left_join(clean_data_table[,.(sample,
                               UNC13A_3prime_leaf,
                               UNC13A_5prime_1_leaf,
                               disease_group2,
                               disease_tissue)], by = 'sample') %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  filter(tissue_clean != "Thoracic_Spinal_Cord") %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-")) %>% 
  add_count(disease_group2,tissue_clean) %>% 
  mutate(n = n / 5) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n {n}")) %>% 
  filter(!is.na(inclusion_reads)) %>% 
  unique() %>% 
  ggplot(aes(x = cell, y = deconv, fill = unc13a_detected)) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.color = NA) + 
  facet_wrap(~plot_name+tissue,scales = 'free') +
  ggpubr::stat_compare_means(label = "p.signif") +
  ggpubr::stat_compare_means(label = "p.format",vjust = 1.2) +
  labs(title = "Cell type decomposition and UNC13A CE detection ",subtitle = "Wilcox test") + 
  ggpubr::theme_pubr()

@@

in_psi = rs12608932_comparisons %>%
  ggplot(aes(y = unc13a_cryptic_leaf_psi, x = detection_name, fill = rs12608932)) +
  geom_boxplot() +
  geom_jitter( height = 0) +
  ylab(element_blank()) +
  xlab(element_blank()) +
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ggpubr::theme_pubr() +
  theme(legend.position = "none") +
  theme(text = element_text(size = 24)) +
  stat_compare_means() +
  stat_compare_means(comparisons = comp_for_gg,label = "p.signif") +
    scale_y_continuous(trans='log10',labels = scales::percent_format(accuracy = 1L))  +
  ggtitle("intronic SNP (rs12608932)")
```

@@

```{r}
pretty_effect_plot = function(model, p_value_cutoff = 0.05, title = "Model 1"){
    p_value = Tmisc::lmp(model)
    r_squared = summary(model)$adj.r.squared
    pretty_plot = broom::tidy(model) %>% 
        filter(p.value < p_value_cutoff) %>%
        mutate(term = fct_reorder(term, -abs(estimate))) %>%
        ggplot(aes(x = term, y = estimate,fill = estimate > 0 )) +
        geom_col() +
        coord_flip() +
        ggtitle(title,subtitle = glue::glue("Model p-value: {p_value} \n Adj. R-squared: {r_squared}"))
}

detect2 = clean_data_table %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(unc13a_cryptic_leaf_psi > 0 ) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(junction_reads_stmn2 >= 30) %>%
  filter(junction_reads_unc13a >= 30) %>%
  filter(disease_tissue == T) %>% 
  mutate(call = fct_relevel(rs12973192,
                              "C/C", "C/G", "G/G")) %>% 
    mutate(number_g_alleles = as.numeric(call)) %>% 
  mutate(age_at_death = as.numeric(age_at_death))

m1 = lm(unc13a_cryptic_leaf_psi ~ stmn_2_cryptic_psi_leaf * 
          number_g_alleles + 
          tissue_clean + 
          age_at_death + 
          sex, 
              detect2)

k = pretty_effect_plot(m1,p_value_cutoff = 0.05)
print(k)

@@

## Log2 Fold UNC / STMN2 CE in TDP-path with both detected - STR - cortex and cord

```{r str and log2 graph}
str_genotype = janitor::clean_names(fread("/Users/annaleigh/Documents/GitHub/unc13a_cryptic_splicing/data/rs56041637.genotypes.377samples.txt")) %>% 
  dplyr::rename(presumed_unique_id = number_sample_id)

detected_table_for_graph = clean_data_table %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(unc13a_cryptic_leaf_psi > 0 ) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(junction_reads_stmn2 >= 30) %>%
  filter(junction_reads_unc13a >= 30) %>%
  filter(disease_tissue == T) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  dplyr::select(sample,participant_id,tissue_clean,disease_group2,rs12973192,rs12608932,stmn_2_cryptic_psi_leaf,unc13a_cryptic_leaf_psi) %>%
  mutate(log2_unc = log2((unc13a_cryptic_leaf_psi)/(stmn_2_cryptic_psi_leaf))) %>% 
  mutate(tissue_clean = gsub("_","\n",tissue_clean)) %>% 
  left_join(str_genotype) %>% 
    filter(!is.na(genotype_ci)) %>% unique()

comparisons = combn(unique(detected_table_for_graph$genotype),2,simplify = F)

detected_table_for_graph %>%
  ggplot(aes(y = log2_unc, x = genotype,fill = genotype)) +
  geom_boxplot(show.legend = F) +
  geom_jitter(show.legend = F, alpha = 0.8, height = 0,pch = 21) +
  stat_compare_means( label = "p.format") +
  stat_compare_means(aes(group = detection_name),comparisons = comparisons,label = "p.signif") +
  ylab("Ratio UNC13A CE PSI / STMN2 PSI") +
  ylab(expression(paste("Lo", g[2], " fold UNC13A CE PSI / STMN2 CE PSI "))) +
  xlab("UNC13A STR Genotype") +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24)) +
  geom_hline(yintercept = 0,size = 1.5)
```

## STR by SNP genotype

```{r}
clean_data_table %>% 
  mutate(presumed_unique_id = gsub("-b38","",participant_id)) %>% 
  select(participant_id,presumed_unique_id,rs12973192,rs12608932) %>% 
  unique() %>% 
  left_join(str_genotype) %>% 
  unique() %>% 
  mutate(genotype = fct_relevel(genotype,"6/6", "6/7", "6/8", "6/9",
                                "6/10","6/12","7/8","8/8","8/10","9/10")) %>% 
  ggplot(aes(x = genotype,fill = rs12973192)) + 
  geom_bar() + 
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ylab("N individuals") +
  xlab("STR genotype")

@@

```{r message=FALSE, warning=FALSE}
read_length_filter <- 29  # length of reads of interest
gencode_v29_info <- fread(paste0(here(), "/data/longest_proteincoding_transcript_hs_details.txt"))
ribo_df <- fread(paste0(here(), "/data/riboseq/sh_tdpb.Aligned.toTranscriptome.out.bam.bed.gz")) 
names(ribo_df)  = c("transcript_id", "start0", "end", "strand")
ribo_df = ribo_df %>% 
    inner_join(gencode_v29_info, by = "transcript_id") %>%
    dplyr::filter(end-start0 == read_length_filter, strand == "+") %>%
    mutate(distance_from_start = start0-cds_start) %>%
    group_by(distance_from_start) %>%
    mutate(n_this_dist = n()) %>%
    dplyr::select(distance_from_start, n_this_dist) %>%
    unique()

@@

```{r als_ftd_genes_venn}
als_ftd_genomics = c("APP", "CHMP2B", "CSF1R", "DNAJC5","UNC13A",
                     "DNMT1", "EPM2A", "GRN", "ITM2B",
                     "MAPT", "NHLRC1", "NOTCH3", "PRNP", 
                     "PSEN1", "PSEN2", "TARDBP", "TBK1", 
                     "TYROBP", "UBQLN2", "VCP", "ATN1", 
                     "ATXN10", "ATXN1", "ATXN2", "C9orf72", 
                     "HTT", "JPH3", "NOP56", "TBP", "ALS2", 
                     "ANG", "DCTN1", "FIG4", "FUS", "HNRNPA1", 
                     "OPTN", "PFN1", "SETX", "SIGMAR1", "SLC52A2", 
                     "SLC52A3", "SOD1", "VAPB", "AR")

ipsc_de = as.data.table(ipsc_de)
ipsc_splicing = as.data.table(ipsc_splicing)

ipsc_de[gene_name %in% als_ftd_risk_genes][order(log2FoldChange)][,.(log2FoldChange,padj,gene_name,gene_expresion)] %>% mutate(has_cryptic = gene_name %in% ipsc_splicing[cryptic_junction == T,gene_name]) %>% unique() 
```
